---
title: "2015 Boston Marathon Results Analysis"
author: "Joshua Gardner -  jpgard.com"
date: "April 5, 2016"
output: html_document
---

#Background and Motivation

The Boston Marathon is one of the world's oldest and most prestigious marathons. Begun in 1897, the Boston Marathon is the world's oldest annual marathon and ranks as one of the world's best-known racing events. It is one of six World Marathon Majors, and is one of four major events held in the United States through the years of both World Wars (Kentucky Derby, Rose Parade, and Westminster Kennel Club Dog Show are the others). The Boston Marathon attracts olympians and world record-holders each year, and as many as 50,000 amateur and professional runners gather to join an elite and competitive field of runners. The Boston Marathon is particularly unique in that registration is not open to all participants--runners must qualify by running a specified marathon time for their age group and gender--making the race especially prestigious and a career-long goal for many runners. Held on Patriot's Day each year, the race attracts well over 500,000 spectators along its historic and famous 26.2 mile course from Hopkinton into downtown Boston.

This project was motivated by several factors. Despite how famous and historic the race and its course are, very little is publicly shared about how participants actually perform in the marathon. The Boston Marathon makes searchable results available on their website, but reports very little about how runners actually perform on the course. Given that this is one of the world's most prestigious marathons, and given the unique population of runners that participate in this race, it seemed like an especially interesting and useful dataset to obtain and explore. Additionally, having participated in the boston marathon five times as a runner myself, I was particularly interested in learning more



#Data Preprocessing

```{r}
library(plyr)
library(lubridate)
library(ggplot2)
results = read.csv("resultsfinal.csv", stringsAsFactors=FALSE)

#create divisions for age groups, based on agr goups listed at http://raceday.baa.org/statistics.html
results$age_group = cut(results$age, breaks = c(0, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80), labels = c("18-34", "35-39", "40-44", "45-49", "50-54", "60-64", "65-69", "70-74", "75-79", "80+"), right = FALSE, ordered_result = TRUE)

#import qualifying times and add to data
qualTimes = read.table("qualifyingTimes.txt", sep = '\t', colClasses = c("factor", "factor", "character"))
names(qualTimes) = c("age_group", "gender", "qualifying_time")

#convert gender, county_origin to factor variables
results$gender = factor(results$gender); results$gender = factor(results$gender); results$county = factor(results$county)

results = merge(results, qualTimes)

#convert times to duration objects using as.difftime() and lubridate's as.duration()

results$X5k <- as.duration(as.difftime(tim = results$X5k, format = "%T", units = "secs"))
results$X10k <- as.duration(as.difftime(tim = results$X10k, format = "%T", units = "secs"))
results$X15k <- as.duration(as.difftime(tim = results$X15k, format = "%T", units = "secs"))
results$X20k <- as.duration(as.difftime(tim = results$X20k, format = "%T", units = "secs"))
results$half <- as.duration(as.difftime(tim = results$half, format = "%T", units = "secs"))
results$X25k <- as.duration(as.difftime(tim = results$X25k, format = "%T", units = "secs"))
results$X30k <- as.duration(as.difftime(tim = results$X30k, format = "%T", units = "secs"))
results$X35k <- as.duration(as.difftime(tim = results$X35k, format = "%T", units = "secs"))
results$X40k <- as.duration(as.difftime(tim = results$X40k, format = "%T", units = "secs"))
results$pace <- as.duration(as.difftime(tim = results$pace, format = "%T", units = "secs"))
results$official_time <- as.duration(as.difftime(tim = results$official_time, format = "%T", units = "secs"))
results$qualifying_time <- as.duration(as.difftime(tim = results$qualifying_time, format = "%T", units = "secs"))

#results$X5k <- as.POSIXlt(results$X5k, format = "%T")
#results$X10k <- as.POSIXlt(results$X10k, format = "%T")
#results$X15k <- as.POSIXlt(results$X15k, format = "%T")
#results$X20k <- as.POSIXlt(results$X20k, format = "%T")
#results$half <- as.POSIXlt(results$half, format = "%T")
#results$X25k <- as.POSIXlt(results$X25k, format = "%T")
#results$X30k <- as.POSIXlt(results$X30k, format = "%T")
#results$X35k <- as.POSIXlt(results$X35k, format = "%T")
#results$X40k <- as.POSIXlt(results$X40k, format = "%T")
#results$pace <- as.POSIXlt(results$pace, format = "%T")
#results$official_time <- as.POSIXlt(results$official_time, format = "%T")

#exclude any columns with na for any value; this removes 194 runners who didn't cross various checkpoints for any reasin (dropped out, missed checkpoint, foot RFID tag error...)
results = na.omit(results)
```

#Analysis: Four Guiding Questions

##1. Who are the runners of the Boston Marathon? 
####Where do most runners come from, what are their ages, what genders are they?

```{r}
countrycounts = data.frame(table(results$county))
names(countrycounts) <- c("Country", "Entrants")
ggplot(countrycounts, aes(x = reorder(Country, -Entrants), y = Entrants)) + geom_bar(stat="identity") + xlab("Country") + ggtitle("Entrants By Country") + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

#TODO: print top 10 sorted

#TODO: print countries with <5 entries just for fun?

#TODO: filled country map
```

Individually visualized, the age distributions of female (top) and male (bottom) runners are shown below:
```{r}
ggplot(results, aes(x=age)) + geom_histogram(fill = "dodgerblue", colour="black", alpha = 0.3) + facet_grid(gender ~ .)
```

We can also directly compare the age distributions of male and female runners by overlaying them:
```{r}
ggplot(results, aes(x=age, fill = gender)) + geom_density(alpha = 0.3)
```
As seen above, females have a relatively higher proportion of runners younger than approximately 45 years of age, and males have a relatively higher proportion of runners older than 45 (although, in general, the distributions are wuite similar).

##2. How fast are the runners of the Boston Marathon? 
####What are the average finishing times for men and women, and within age groups? How do these finishing times compare to the qualifying times needed to gain entry into the race?

```{r}

ggplot(results, aes(x=age_group, y=official_time)) + geom_violin(fill = "dodgerblue", alpha = 0.3) + ggtitle("Official Finishing Times By Age Group")
ggplot(results, aes(x=age_group, y=official_time)) + geom_boxplot(fill = "dodgerblue", alpha = 0.3) + ggtitle("Official Finishing Times By Age Group")

ggplot(results, aes(x=gender, y=official_time)) + geom_violin(fill = "dodgerblue", alpha = 0.3) + ggtitle("Official Finishing Times By Age Group")
ggplot(results, aes(x=gender, y=official_time)) + geom_boxplot(fill = "dodgerblue", alpha = 0.3) + ggtitle("Official Finishing Times By Age Group")

#faceted plots with qualifying times for men, women
ggplot(subset(results, gender == "M"), aes(x=gender, y=official_time)) + geom_violin(fill = "dodgerblue", alpha = 0.5) + ggtitle("Official Finishing Times By Age Group (Men)\n Relative to Qualifying Standard")  + geom_hline(aes(yintercept=qualifying_time), size=1, colour = "gold") + facet_grid(. ~ age_group) + theme(axis.text.x = element_blank(), axis.title.x=element_blank(), axis.title.y=element_blank()) + scale_y_continuous(breaks=c(3600, 7200, 10800, 14400, 18000, 21600), labels=c("1:00:00","2:00:00", "3:00:00", "4:00:00", "5:00:00", "6:00:00"))


ggplot(subset(results, gender == "F"), aes(x=gender, y=official_time)) + geom_violin(fill = "dodgerblue", alpha = 0.5) + ggtitle("Official Finishing Times By Age Group (Women)\n Relative to Qualifying Standard")  + geom_hline(aes(yintercept=qualifying_time), size=1, colour = "gold") + facet_grid(. ~ age_group) + theme(axis.text.x = element_blank(), axis.title.x=element_blank(), axis.title.y=element_blank()) + scale_y_continuous(breaks=c(3600, 7200, 10800, 14400, 18000, 21600), labels=c("1:00:00","2:00:00", "3:00:00", "4:00:00", "5:00:00", "6:00:00"))


#TODO: split violin plots for male/female, within each age group --  see http://stackoverflow.com/questions/35717353/split-violin-plot-with-ggplot2
```


##3. How do runners run the race? 
####Do they typically start fast, and slow down as the race progresses and they become fatigued, or do the more experienced runners that qualify for Boston pace themselves and achieve an even performance? How does this vary across age groups and genders?

```{r}
#create variable for second half
results$sec_half = results$official_time - results$half


half_times <- results[,c("bib_number", "gender", "age_group", "half", "sec_half")] %>%
  gather(half, time, half:sec_half)
 
half_times$half = factor(ifelse(half_times$half == "half", 1, 2))
half_times$time = as.numeric(half_times$time)

ggplot(half_times, aes(x = half, y = time)) + geom_violin(fill = "dodgerblue", alpha = 0.5)

#Main plot--TODO: adjust color
ggplot(half_times, aes(x = half, y = time)) + geom_violin(aes(fill = half, colour = half), alpha = 0.5) + facet_grid(gender ~ age_group) + stat_summary(fun.y=median, geom="point", fill="white", shape=21, size=2.5)

ggplot(half_times, aes(x = half, y = time)) + geom_jitter(aes(colour = gender, alpha = 0.5))


#plot difference between first 5k and last 5k
ggplot(results, aes(x=age_group, y=results$X40k - results$X35k-results$X5k)) + geom_violin(fill = "dodgerblue", alpha = 0.3) + stat_summary(fun.y=median, geom="point", fill="gold", shape=21, size=2.5)

ggplot(results, aes(x=age_group, y=results$X40k - results$X35k-results$X5k)) + geom_jitter(colour = "dodgerblue", alpha = 0.3) + scale_y_continuous(limits = c(-1800, 1800), breaks = c(-1800, -1200, -600, 0, 600, 1200, 1800), labels = c("-30:00", "-20:00", "-10:00", "00:00", "+10:00", "+20:00", "+30:00")) + ylab("Difference in minutes") + ggtitle("Difference between first and final 5k times\nby age group")
```

```{r}
#Data processing for pace calculation along various 5k segments
library(dplyr)
segment_5k_times <- results %>%
    mutate(segment_5k = X5k, segment_10k = X10k - X5k, segment_15k = X15k-X10k, segment_20k = X20k-X15k, segment_25k = X25k-X20k, segment_30k = X30k-X25k, segment_35k=X35k-X30k, segment_40k = X40k-X35k) %>%
    gather(km_segment, time, segment_5k:segment_40k) %>%
    mutate(km_segment = as.factor(as.numeric(gsub("k", "", gsub("segment_", "", km_segment))))) %>%
    subset(select = -c(name, age, bib_number, city, origin, X5k:sec_half_pace))

ggplot(filter(segment_5k_times, age_group <= "45-49"), aes(x = km_segment, y = time)) + geom_jitter(colour = "dodgerblue", alpha = 0.3)+ geom_boxplot(fill = "white", width = 0.2, outlier.shape = NA) + facet_grid(age_group ~ gender) + xlab("5k segment") + ylab("Segment Time") + scale_y_continuous(limits = c(900, 2700), breaks = c(1200,  1800,  2400), labels = c("20:00", "30:00", "40:00")) + ggtitle("Performance over 5k race segments by gender and age group\n2015 Boston Marathon")

ggplot(filter(segment_5k_times, age_group > "45-49"), aes(x = km_segment, y = time)) + geom_jitter(colour = "dodgerblue", alpha = 0.3)+ geom_boxplot(fill = "white", width = 0.2, outlier.shape = NA) + facet_grid(age_group ~ gender) + xlab("5k segment") + ylab("Segment Time") + scale_y_continuous(limits = c(900, 2700), breaks = c(1200,  1800,  2400), labels = c("20:00", "30:00", "40:00")) + ggtitle("Performance over 5k race segments by gender and age group\n2015 Boston Marathon")
    
```

##4. Who doesn’t finish the race? 
####Each year, a small number of runners (approximately 1,000) do not finish the race, which results in a result typically referred to as ‘DNF’ (Did Not Finish). What types of runners comprise this group, in terms of age, gender, and nationality?

#Appendix: Extra Plots and Exploratory Analysis

#### How is pace related to age group?
```{r}
ggplot(results[results$age_group!="NA",], aes(x=age_group, y=pace)) + geom_violin()
```

####How is bib number (= rank based on qualifying time) related to overall performance? Do runners with faster qualifying times place higher?
```{r}
#note: these are interesting graphs but need a little more wrangling to come out right!

#ggplot(results, aes(x = bib_number, y = overall_place)) + geom_point()
#ggplot(results, aes(x = bib_number, y = overall_place)) + stat_binhex()
```